TacticToe: Learning the Game of Tactics

### 2018-01-11 tactictoe_jv.tex
figure for search tree

\todo{These 2 paragraphs are copied}
Many of the state-of-the-art interactive theorem provers (ITPs) such as 
  \holfour~\cite{hol4}, \hollight~\cite{Harrison09hollight}, 
  \isabelle~\cite{isabelle} 
  and \coq~\cite{coq-book}
  provide high-level parameterizable tactics for constructing the
  proofs.  Such tactics typically analyze the current goal state and
  assumptions, apply nontrivial proof transformations, which get
  expanded into possibly many basic kernel-level inferences or significant 
  parts of the proof term.
%\marginpar{not Coq}
  In this
  work we develop a tactic-level automation procedure for the \holfour ITP
  which guides selection of the tactics by learning from previous
  proofs.  Instead of relying on translation to first-order automated
  theorem provers (ATPs) as done by the hammer 
  systems~\cite{hammers4qed,tgck-cpp15}, the technique
%\marginpar{too many cites?}
%BlanchetteGKKU16,holyhammer,mizAR40
  directly searches for sequences of tactic applications that lead to
  the ITP proof, thus avoiding the translation and
  proof-reconstruction phases needed by the hammers.

@@

\begin{itemize}
\item Proof recording at the tactic level is made more precise. Pattern 
matching constructions, opening of submodules are supported. The different 
recording steps (parsing, unfolding and replaying) are regrouped into a single 
\sml function. Intermediate files storing the result of each step are not 
necessary anymore. So, our recording is more complete and fully 
automatized.
\item Existing tactic orthogonalization on recorded data is made stronger by 
also including the internal ATP \metis in the orthogonalization 
competition. Because we manage a database of theorems for argument 
prediction, we apply here a pruning step to remove easy theorems.
\item Two types of arguments (list of theorems and terms) can be predicted 
independently. Combinatorial explosion in the number of tactics 
is prevented through the introduction of abstracted tactics. As desired, their 
instantiations is dependent on the target goal.
\item MCTS (Monte Carlo tree search) now replaces A* in our proof search 
algorithm. MCTS gives a more balanced feedback to guide proof search by 
comparing subtrees instead of leafs. (todo: A first successful evaluation 
function.)
\item The internal ATP \metis used during our proof search is complemented by 
calls to \holyhammer. For efficiency reasons, those calls are run in parallel 
and asynchronously.
\item First evaluation of \tactictoe and \holyhammer success rates relative to 
the length of the original proofs.
\item User interaction is enhanced by the minimization and prettification 
applied on the generated proof.
\end{itemize}

@@

\subsection{Dependencies}\label{sec:dependencies}
We first rank feature vectors by their distance to the conjecture. But better 
ranked vectors only contains symbols from 
the conjecture and may not be suitable for later stages of the proof where the 
goals may have derived significantly from the conjecture. To solve this 
problem, we reorder the list so that descendants (see 
Definition~\ref{def:descendant}) of vectors inherits the similarity of their 
parent.

@@

We augment the ordered predictions by inserting after each predicted vector 
its descendants and removing lower ranked duplicates.

This is algorithm is used to predict tactics in the preselection
phase of the algorithm. Similarly, dependencies between theorems are used to 
preselect theorems.

\subsection{Preselection}\label{sec:pretactic}
Since predicting efficiently is crucial during proof 
search, we preselect a number of tactics before the search.
We take the first 500 vectors of the augmented ordered list. Only members of 
this list can be predicted during proof search. When an 
open goal $g$ is created, the scores of those 500 preselected tactics are
recomputed which gives a preferential order of tactics to be tried on $g$.

@@

\begin{definition} (Coverage)\\ 
Let $\mathfrak{F}$ be a recorded set of feature vectors. We can define the 
coverage $C(t)$ of a tactic $t$ by the number of times this tactic appears in 
$\mathfrak{F}$. Expressing this in a formula, we get:
  \[C(t) =_{def} card\ \lbrace (x,g)\ |\ (x,g) \in \mathfrak{F} \wedge x = t
  \rbrace  \]
Intuitively, this notion estimates how general a tactic is by counting the 
number of different goals it is useful for.
\end{definition}

\begin{definition} (Goal subsumption)\\ 
A goal subsumption $S$ is a partial reflexive relation on a set of goals.\\
Given two goals $g_1$ and $g_2$, we note:
$g_1 \ge g_2 \Leftrightarrow_{def} (g_1,g_2) \in S$.\\
Assuming we have a way to test if a goal is a consequence of another, a 
possibly useful subsumption definition would be: 
$g_1 \ge g_2  \Leftrightarrow_{def} g_1\ \mbox{is a consequence of}\ g_2$\\
In our experiments, we take the minimal subsumption defined by $g_1 \ge g_2  
\Leftrightarrow_{def} g_1 =_{\alpha} g_2$ where $=_{\alpha}$ is 
$\alpha$-equivalence.
\end{definition}

We can naturally extend a goal subsumption to a set of goals in the following 
way.
\begin{definition} (Goal set subsumption)\\
A goal set subsumption is an partial relation on a set of goals.
Given two sets of goals $l_1$ and $l_2$, we can define it from goal 
subsumption $\ge$ by:
\[l_1 \ge l_2  \Leftrightarrow_{def} \forall g_2 \in l_2\ \exists g_1 \in l_1\ 
g_1 \ge g_2\]
\end{definition}

@@

We can now express the orthogonalization algorithm given in a single formula 
with parameters: $P_n$ for the tactic predictor where $n$ for the number of 
predicted
tactics, $S$ for the tactic subsumption relation and $C$ for the coverage of 
a tactic. The orthogonalization process $\mathbb{O}$ is given as input a 
recorded feature vector $(t,g)$ where $t$ is the tactic originally applied to 
the goal $g$. The winning tactic of the competition is given by the formula:

\[\mathbb{O}(t,g) = \argmax_{x\ \in\ P_n(g) \cup \lbrace t \rbrace} \lbrace 
C(x)\ 
|\ x \ge_g t\rbrace\]

After orthogonalization, because the orthogonalized tactic is supposedly 
better, we record the feature vector $(\mathbb{O}(t,g),g)$ instead of 
$(t,g)$. To sum up, the winning tactic is the one with the most coverage and 
that is able to replicate the effect of the original tactic (or do better if we 
woudl consider more complex goal subsumptions).

@@

\section{Proof Search}\label{sec:proofsearch}
Despite the best efforts of the prediction algorithms, the selected tactic may 
not 
solve the current goal, proceed in the wrong direction
or even loop. For that reason, the prediction needs to be accompanied by a  
proof search mechanism that allows for backtracking and 
can choose which proof tree to extend next and in 
which direction.
Our search algorithm takes inspiration from advanced tree search algorithm such 
that

@@

\subsection{Diagonalized tree search}
(Move this section to Monte Carlo tree search and simplify it) 
(Don't mention diagonlazied tree search)

@@

\subsection{Monte Carlo Tree Search}

We will reuse the prediction

@@

@@

The third point is a partial attempt to prevent confluent branches. The general 
case where two branches becomes parallel after $n$ steps is handle by a
tactic cache which memorizes tactic effects on goals. Similarly there is also a 
prediction cache, which memorize the tactic predictions for each goal. Those 
two caches allow a fast exploration of confluent branches but they remain 
separated in our search tree.
\end{definition}

@@

@@

\begin{table*}[t]
\centering\ra{1.3}
\small
\begin{tabular}{llllll}
\toprule
 ID & \multicolumn{2}{c}{nodes}   & \multicolumn{2}{c}{proof size} & 
 \multicolumn{1}{c}{time}\\ \cmidrule(lr){2-3} \cmidrule(lr){4-5} 
 \cmidrule(lr){6-6}
     & average & max & average & max & average \\
\midrule
 \tactictoe & 94.66 & 421 & 3.34 & 39 & 0.66 \\
\bottomrule
\end{tabular}
 \caption{\label{tab:stats}search statistics}
\end{table*}

@@

\section{Case study}
Investigating further the different qualities of \tactictoe, we study its 
generated proof scripts on an example in \texttt{list} theory (see 
Example~\ref{ex:casestudy}). 
The theorem to be proven states the equivalence 
between the fact that a number $n$ is greater than the length of a list 
$\mathit{ls}$ with the 
fact that dropping $n$ elements from this list returns an empty list.

The human proof proceeds by induction on $n$ followed by solving both goals 
using rewrite steps combined with an arithmetic decision procedure. 
Both \tactictoe proofs (NH and SH) follow the general idea of reasoning by 
induction but solve 
the base case and the inductive case in a different way.
The base case only needs rewriting using the global simpset in the 
\tactictoe(NH)
proof, which is simulated by a call to \metis in the (SH) proof.
The inductive case should require an arithmetic decision procedure as hinted by 
the human 
proof. This is achieved by rewriting using an arithmetic simpset in the second 
proof. In the first proof however, a rewriting step and case splitting step 
were used
to arrive at a point where \metis calls succeed.
The tactic proof produced by \tactictoe(NH) often looks better than the one 
discovered by \tactictoe(SH) in that it does not involve \metis calls with a 
large numbers of premises.

>>>>

\todo{These 2 paragraphs are copied and slightly edited}
Many of the state-of-the-art interactive theorem provers (ITPs) such as 
  \holfour~\cite{hol4}, \hollight~\cite{Harrison09hollight}, 
  \isabelle~\cite{isabelle} 
  and \coq~\cite{coq-book}
  provide high-level parameterizable tactics for constructing the
  proofs.  Such tactics typically analyze the current goal state and
  assumptions, apply nontrivial proof transformations, which get
  expanded into possibly many basic kernel-level inferences or significant 
  parts of the proof term.
%\marginpar{not Coq}
  In this
  work we develop a tactic-level automation procedure for the \holfour ITP
  which guides selection of the tactics by learning from previous
  proofs.  Instead of relying on translation to first-order automated
  theorem provers (ATPs) as done by the hammer 
  systems~\cite{hammers4qed,tgck-cpp15}, the technique
%\marginpar{too many cites?}
%BlanchetteGKKU16,holyhammer,mizAR40
  directly searches for sequences of tactic applications that lead to
  the ITP proof, thus avoiding the translation and
  proof-reconstruction phases needed by the hammers.

@@

\begin{itemize}
\item Proof recording at the tactic level is made more precise. Pattern 
matching constructions, opening of submodules are supported. The different 
recording steps (parsing, unfolding and replaying) are regrouped into a single 
\sml function. Intermediate files storing the result of each step are not 
necessary anymore. So, our recording is more complete and fully 
automatized.
\item Existing tactic orthogonalization on recorded data is made stronger by 
also including the internal ATP \metis in the orthogonalization 
competition. Because we manage a database of theorems for argument 
prediction, we apply here a pruning step to remove easy theorems.
\item Two types of arguments (list of theorems and terms) can be predicted 
independently. Combinatorial explosion in the number of tactics 
is prevented through the introduction of abstracted tactics. As desired, their 
instantiations is dependent on the target goal.
\item MCTS (Monte Carlo tree search) now replaces A* in our proof search 
algorithm. MCTS gives a more balanced feedback to guide proof search by 
comparing subtrees instead of leafs. (todo: A first successful evaluation 
function.)
\item The internal ATP \metis used during our proof search is complemented by 
asynchronous calls to \eprover.
\item First evaluation of \tactictoe and \eprover success rates relative to 
the length of the original proofs.
\item User interaction is enhanced by the minimization and prettification 
applied on the generated proof.
\end{itemize}

@@

\subsection{Preselecting feature vectors and their 
dependencies}\label{sec:dependencies}
We first rank feature vectors by their distance to the conjecture. But better 
ranked vectors only contains symbols from 
the conjecture and may not be suitable for later stages of the proof where the 
goals may have derived significantly from the conjecture. To solve this 
problem, we reorder the list so that descendants (see 
Definition~\ref{def:descendant}) of vectors inherits the similarity of their 
parent.

@@

We augment the ordered predictions by inserting after each predicted vector 
its descendants and removing lower ranked duplicates. This contrast with 
theorem predictions where we only select parents. This phase is only applied 
during preselection, so is of no great consequence as long as we select enough 
theorems and tactics.

Since predicting efficiently is crucial during proof 
search, we preselect a number of tactics before the search.
We take the first 500 vectors of the augmented ordered list. Only members of 
this list can be predicted during proof search. When an 
open goal $g$ is created, the scores of those 500 preselected tactics are
recomputed which gives a preferential order of tactics to be tried on $g$.

@@

\begin{definition} (Coverage)\\ 
Let $\mathfrak{F}$ be a recorded set of feature vectors. We can define the 
coverage $\textit{\mbox{Coverage}}(t)$ of a tactic $t$ by the number of times 
this tactic 
appears in 
$\mathfrak{F}$. Expressing this in a formula, we get:
  \[\textit{\mbox{Coverage}}(t) =_{def} card\ \lbrace (x,g)\ |\ (x,g) \in 
  \mathfrak{F} 
  \wedge x 
  = t
  \rbrace  \]
Intuitively, this notion estimates how general a tactic is by counting the 
number of different goals it is useful for.
\end{definition}

\begin{definition} (Goal subsumption)\\ 
A goal subsumption $S$ is a partial relation on a set of goals.\\
Given two goals $g_1$ and $g_2$, we note $g_1 \ge g_2$ when  $(g_1,g_2) \in 
S$.\\
Assuming we have a way to test if a goal is a consequence of another, a 
possibly useful subsumption definition would be: 
\[g_1 \ge g_2  \Leftrightarrow_{def} g_1 \mbox{ is a consequence of } g_2\]
In our experiments, we take a minimal subsumption defined by:
\[g_1 \ge g_2  \Leftrightarrow_{def} g_1 \mbox{ is $\alpha$-equivalent to }  
 g_2\]
\end{definition}

We can naturally extend a goal subsumption to a set of goals.
\begin{definition} (Goal set subsumption)\\
A goal set subsumption is an partial relation on a set of goals.
Given two sets of goals $l_1$ and $l_2$, we can define it from goal 
subsumption $\ge$ by:
\[l_1 \ge l_2  \Leftrightarrow_{def} \forall g_2 \in l_2\ \exists g_1 \in l_1\ 
g_1 \ge g_2\]
\end{definition}

@@

We can now express the orthogonalization algorithm given in a single formula 
with parameters: $P_n$ for the tactic predictor where $n$ for the number of 
predicted
tactics, $S$ for the tactic subsumption relation and $\textit{\mbox{Coverage}}$ 
for the coverage of 
a tactic. The orthogonalization process $\mathbb{O}$ is given as input a 
recorded feature vector $(t,g)$ where $t$ is the tactic originally applied to 
the goal $g$. The winning tactic of the competition is given by the formula:

\[\mathbb{O}(t,g) = \argmax_{x\ \in\ P_n(g) \cup \lbrace t \rbrace} \lbrace 
\textit{\mbox{Coverage}}(x)\ 
|\ x \ge_g t\rbrace\]

After orthogonalization, because the orthogonalized tactic is supposedly 
better, we record the feature vector $(\mathbb{O}(t,g),g)$ instead of 
$(t,g)$. To sum up, the winning tactic is the one with the most coverage and 
that is able to replicate the effect of the original tactic.

@@

\section{Proof Search}\label{sec:proofsearch}
Despite the best efforts of the prediction algorithms, the selected tactic may 
not 
solve the current goal, proceed in the wrong direction
or even loop. For that reason, the prediction needs to be accompanied by a  
proof search mechanism that allows for backtracking and 
can choose which proof tree to extend next and in 
which direction.
Our search algorithm is an advanced tree search algorithm that relies on 
evaluation and policy functions to guide the search.

@@

%\subsection{Diagonalized tree search}
%(Move this section to Monte Carlo tree search and simplify it) 
%(Don't mention diagonlazied tree search)

@@

@@

\subsection{A proof tree and solved node}

A tactical proof tree is a set of nodes containing a list of goals where each 
goal $g$ of a node $n$ is linked to a node $m$ when the list of

\begin{figure}{}
\begin{center}
\begin{tikzpicture}[node distance=0.7cm]
\node [tcircle] (0) {$g_0^i$};
\node [right of=0] (0r) {.\ .\ .};
\node [tcircle,right of=0r] (0rr) {$g_0^n$};
\node [left of=0] (0l) {.\ .\ .};
\node [tcircle,left of=0l] (0ll) {$g_0^1$};
\node [draw,fit=(0rr) (0ll)] {}; 
\node [above of=0,node distance=2cm] (2) {.\ .\ .};
\node [left of=2, tcircle] (2l) {$g_2^1$};
\node [tcircle,right of=2] (2r) {$g_2^p$};
\node [draw, fit=(2l) (2r)] (2b) {};

\node [tcircle, left of=2l, node distance=2cm] (1r) {$g_1^q$};
\node [left of=1r] (1) {.\ .\ .};
\node [tcircle, left of=1] (1l) {$g_1^1$};
\node [draw, fit=(1l) (1r)] (1b) {};

\node [right of=2r,node distance=2cm] (3l) {$\phantom{g_3^p}$};
\node [right of=3l] (3) {\phantom{.\ .\ .}};
\node [right of=3] (3r) {$\phantom{g_3^p}$};
\node [fit=(3l) (3r)] (3b) {};

\draw[-to,thick] (0) to node[xshift=-10](t1){$t_1$} (1b);
\draw[-to,thick] (0) to node[xshift=-10](t2){$t_2$} (2b);
\draw[-to,dotted,thick] (0) to (3b);
\end{tikzpicture}
\end{center}
\caption{Choice of a goal and a tactic}
\label{fig:choice}
\end{figure}

Starting from the bottom node, we can construct descendant of this node by 
choosing a goal $g_i^0$ on this node and choosing tactics $t1,t2,\ldots$ to 
apply to this goal. Here the result produces new nodes each containing a list 
of goals to be proven. If one of those list where empty, the goal $g_i^0$ would 
be proved.

A node is solved when every of its goal are solved.
A goal is solved when every of its children node are solved.

\subsection{Guiding the proof search}

@@

\end{definition}

The third point of the definition is a partial attempt to prevent confluent 
branches. The general 
case where two branches becomes parallel after $n$ steps is handle by a
tactic cache which memorizes tactic effects on goals. Similarly there is also a 
prediction cache, which memorize the tactic predictions for each goal. Those 
two caches allow a fast exploration of confluent branches but they remain 
separated in our search tree.

@@

Disclaimer:
Comparing results from different tables is possible but not advisable since 
conditions of the experiments may vary slightly because of various factors such 
as overhead cost of additional features or the fact that they were run on 
different machines on a local machine or on the server.

@@

\begin{table*}[t]
\centering\ra{1.3}
\small
\begin{tabular}{llllll}
\toprule
 ID & \multicolumn{2}{c}{nodes}   & \multicolumn{2}{c}{proof size} & 
 \multicolumn{1}{c}{time}\\ \cmidrule(lr){2-3} \cmidrule(lr){4-5} 
 \cmidrule(lr){6-6}
     & average & max & average & max & average \\
\midrule
 \tactictoe & 94.66 & 421 & 3.34 & 39 & 0.66 \\
\bottomrule
\end{tabular}
 \caption{\label{tab:stats} old search statistics}
\end{table*}

@@

\section{Case study}
%Investigating further the different qualities of \tactictoe, we study its 
%generated proof scripts on an example in \texttt{list} theory (see 
%Example~\ref{ex:casestudy}). 
%The theorem to be proven states the equivalence 
%between the fact that a number $n$ is greater than the length of a list 
%$\mathit{ls}$ with the 
%fact that dropping $n$ elements from this list returns an empty list.
%
%The human proof proceeds by induction on $n$ followed by solving both goals 
%using rewrite steps combined with an arithmetic decision procedure. 
%Both \tactictoe proofs (NH and SH) follow the general idea of reasoning by 
%induction but solve 
%the base case and the inductive case in a different way.
%The base case only needs rewriting using the global simpset in the 
%\tactictoe(NH)
%proof, which is simulated by a call to \metis in the (SH) proof.
%The inductive case should require an arithmetic decision procedure as hinted 
%by 
%the human 
%proof. This is achieved by rewriting using an arithmetic simpset in the second 
%proof. In the first proof however, a rewriting step and case splitting step 
%were used
%to arrive at a point where \metis calls succeed.
%The tactic proof produced by \tactictoe(NH) often looks better than the one 
%discovered by \tactictoe(SH) in that it does not involve \metis calls with a 
%large numbers of premises.
